---
title: "HSC division patterns"
author: "PS Stumpf"
date: ""
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, auto.dep=TRUE,
                      echo=TRUE, tidy=TRUE, width.cutoff=50, fig.width = 4, fig.asp = 1)

suppressPackageStartupMessages({
  library(affy)
  library(limma)
  library(gplots)
  library(marray)
  library(biomaRt)
  library(ggplot2)
  library(dendextend)
  library(dendroextras)
  library(RColorBrewer)
  # library(scatterplot3d)
  })
```



# Ensemble-cell Arrays

__Platform__:

Affymetrix 430.2

__Samples__:

* 3x LSK Evi1-GFP-positive
* 3x LSK Evi1-GFP-negative
* 3x LSK CD34+
LSK is parent population, split into two sub-populations based on Evi1-GFP reporter expression.

## Load normalised ensemble-cell microarray data

Read normalised expression values from tab delimited file and subset columns to relevant values for further analysis.

```{r LoadEnsembleCellArrays, echo=TRUE}
# read tab delimited file
Evi1MAS5 <- read.delim('../Data/Affymetrix_Microarrays/M0133_SigalData_MAS5.TXT')
  # find columns for signal
  SignalColumns <- grepl('.Signal', colnames(Evi1MAS5))

# rename columns using short name  
Evi1MAexprs <- Evi1MAS5[, SignalColumns]
  colnames(Evi1MAexprs) <- c('GP','GP','GP','GN','GN','GN','L','L','L')
# assign group colors
col.line <- c(rep('#BA2025',3), rep('#0072BD',3), rep('#505050',3))

```

## Identifier mapping

Download Affymetrix Mouse430_2 Annotation file from:

 http://www.affymetrix.com/Auth/analysis/downloads/na36/ivt/Mouse430_2.na36.annot.csv.zip

and unzip.

```{r MapAffymetrixRelease36, echo=TRUE, eval=FALSE}
# Lookup-table based on affymetrix' own release (170 MB .csv)
Mouse430.2 <- read.csv('../Data/Affymetrix_Microarrays/Mouse430_2/Mouse430_2.na36.annot.csv', skip = 22)
```

## Principal Component Analysis

Clustering of samples into 3 groups is supported by principal component analysis, and in particular PC1 and PC3.

```{r PrincipalComponents, echo=TRUE}
Evi1MAprcomp <- prcomp(t(log(Evi1MAexprs)))
rownames(Evi1MAprcomp$x) <-  c('Evi1+','Evi1+','Evi1+','Evi1-','Evi1-','Evi1-','CD34+','CD34+','CD34+')

# Variance explained
Evi1MAprcomp$varexpl <-Evi1MAprcomp$sdev^2/sum(Evi1MAprcomp$sdev^2)
```


## Sample relationships in PC1, PC2 and PC3

Three-dimensional scatterplot in principal component space shows clear differentiation between parent population and sub-populations.

Hierarchical clustering on de-noised data (PC1, PC2, PC3) is used to quantify the similarity between clusters (euclidean distance and ward linkage).

```{r PCA3D, echo=TRUE}
# Plot results of PCA in 3D
scatterplot3d::scatterplot3d(Evi1MAprcomp$x[,c(1,3,2)], las=1, color = col.line, pch=19, type='h', xlim= c(-150, 200), ylim = c(-150,150), zlim=c(-150,150), cex.symbols=1.3, box=T, angle=35, mar=c(5,4,2,2), main='Principal component analysis')

# Calculate distance in PCA space (PC1, PC3, PC2)
# ward clustering gives variability within clusters, average clustering uses true distance between average of samples
 par(mar=c(5,4,2,3))
plot(as.dendrogram(hclust(dist(Evi1MAprcomp$x[,1:3], 'euclidean'), 'ward.D')), horiz=T, main='Clustering in PC1, PC2 and PC3')
```


## Differential Gene Expression Analysis using limma

```{r DifferentialExpressionLimma, echo=TRUE}
# using limma
design   <- model.matrix(~0+colnames(Evi1MAexprs))
contrast <- makeContrasts(GP - GN, GP - L, GN - L, levels=c('GN', 'GP', 'L'))
# contrast <- makeContrasts(GP - GN, levels=c('GN', 'GP', 'L')) # optional # for GP-GN contrast only
colnames(design) <- c('GN','GP','L')
fit <- lmFit(log(Evi1MAexprs), design)
fit <- contrasts.fit(fit, contrasts = contrast)
fit <- eBayes(fit)

# plot results and overlap between contrasts - decision threshold based on corrected p-value
fit.result <- decideTests(fit, p.value = 0.05, method='global', adjust.method = 'BH')
vennDiagram(fit.result)

# extract rows of significant genes
Evi1MAlimma <- list()
Evi1MAlimma$features <- list('GP - GN' = which(fit.result[,1] != 0),
                             'GP - L'  = which(fit.result[,2] != 0),
                             'GN - L'  = which(fit.result[,3] != 0))
# Evi1MAlimma$features <- list('GP - GN' = which(fit.result[,1] != 0)) # optional # for GP-GN contrast only

Evi1MAlimma$featexpr <- log(Evi1MAexprs[unique(unlist(Evi1MAlimma$features)),])
rownames(Evi1MAlimma$featexpr) <- Evi1MAS5[rownames(Evi1MAlimma$featexpr) ,1]
```


These differences are visualised using a heatmap of the (feature-wise) z-transformed expression data.

```{r LimmaDendrogram, echo=TRUE, eval=FALSE}
# scale data by feature
Evi1MAlimma$Scaledfeatexpr <- Evi1MAlimma$featexpr
Evi1MAlimma$Scaledfeatexpr[,] <- NA
for (i in 1:nrow(Evi1MAlimma$featexpr)){
   Evi1MAlimma$Scaledfeatexpr[i,] <- as.numeric(scale(as.numeric(Evi1MAlimma$featexpr[i,])))
}

# cluster features
dend.features <- dist(Evi1MAlimma$Scaledfeatexpr, method='euclidean')
dend.features <- as.dendrogram(hclust(dend.features, method='average'))
# dend.features <- dendextend::click_rotate(dend.features, continue=T,las=2)
# dend.features2 <- dendroextras::color_clusters(dend.features, k=2, col = brewer.pal('Dark2', n = 2))

# extract affy ids
a <- names(which(slice(dend.features, k=5)==1))

# cluster samples
dend.specimen <- dist(t(Evi1MAlimma$Scaledfeatexpr), method='euclidean')
dend.specimen <- as.dendrogram(hclust(dend.specimen, method='average'))
# dend.specimen <- dendextend::click_rotate(dend.specimen, continue=T,las=2)
```


```{r LimmaHeatmap, echo=TRUE, eval=TRUE, fig.asp=1/3, fig.height=2, fig.width=6}
# plot heatmap
gplots::heatmap.2(as.matrix(t(Evi1MAlimma$Scaledfeatexpr)), Rowv = dend.specimen, Colv = dend.features,
          RowSideColors = col.line[1:9], colRow = col.line[1:6], scale='n',
          cexRow = 1.1, labCol = FALSE, cexCol = 1.1,
          offsetRow = 0, margins = c(0.5,3.5),
          col = brewer.pal(9,'PRGn'), symbreaks = F, symkey = F,
          trace = "none", density.info = "none",
          lwid = c(.5, 4), lhei = c(1, 4),
          key = T, keysize = 1, key.title = NA, key.xlab = NA,
			    key.par = list(mgp = c(.5,0.5,0), mar = c(2,.5,.5,.5)+.1, cex=.5))
```

